---
title: "FAOSTAT–MODIS data-integration report"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
---

## Purpose and scope

This report audits the integration of the FAOSTAT maize country–year panel
with annual MODIS/Terra net primary productivity (NPP) at one teaching
reference pixel per country.

The reference pixel demonstrates acquisition, spatial metadata, identifier
alignment, temporal overlap, unit scaling, a one-to-one join, and lineage. It
does **not** measure national vegetation, national cropland, or maize fields.
It must not be interpreted as a country-level environmental predictor.

```{r}
#| label: setup
library(dplyr)
library(here)
library(readr)

integrated_file <- here("data-interim", "maize-yield-with-npp.csv")
audit_file <- here("data-processed", "data-integration-audit.csv")
satellite_file <- here("data-raw", "modis-npp-reference-sites.csv")

required_files <- c(integrated_file, audit_file, satellite_file)
missing_files <- required_files[!file.exists(required_files)]

if (length(missing_files) > 0) {
  stop("Run Rscript scripts/integrate-data.R before rendering this report.")
}

integrated <- read_csv(integrated_file, show_col_types = FALSE)
audit <- read_csv(audit_file, show_col_types = FALSE)
satellite <- read_csv(satellite_file, show_col_types = FALSE)
```

## Intended grain, key, and relationship

- FAOSTAT prepared grain: one row per country and calendar year.
- MODIS teaching grain: one row per project country reference pixel and year.
- Integrated grain: one row per project country and calendar year, retaining
  the maize study population.
- Integrated key: `project_country_id + year`.
- Expected join relationship: one-to-one for overlapping keys.
- Join type: left join from the FAOSTAT maize panel.

The two inputs are technically aligned to country identifier and calendar
year. They are not spatially equivalent: a country statistic and one raster
pixel represent very different areas.

## Sources and lineage

The complete source records are in `metadata/source-register.yml`; output-field
lineage is in `metadata/integration-data-dictionary.csv`.

The MODIS teaching snapshot contains the raw integer `Npp_500m` value and
`Npp_QC_500m` percentage returned by the ORNL DAAC service. Integration applies
the documented MOD17A3HGF scale factor of `0.0001` to obtain
kg C/m²/year. Provider fill values from 32761 through 32767 become missing.

Country mappings and WGS84 reference coordinates are explicit in
`metadata/country-crosswalk.csv`. The coordinates identify teaching pixels;
they were not selected from a crop mask or sampling design.

## Join audit

```{r}
#| label: audit-results
audit
```

```{r}
#| label: output-key-check
tibble(
  rows = nrow(integrated),
  unique_keys = n_distinct(integrated$project_country_id, integrated$year),
  countries = n_distinct(integrated$project_country_id),
  first_year = min(integrated$year),
  last_year = max(integrated$year),
  rows_with_satellite = sum(!is.na(integrated$modis_date))
)
```

## Temporal coverage and unmatched keys

The maize panel covers 1990–2022. The compact satellite snapshot deliberately
covers only 2018–2022. Earlier maize keys remain in the integrated output with
missing satellite fields because the left join preserves the maize study
population.

```{r}
#| label: coverage
integrated |>
  mutate(has_satellite_record = !is.na(modis_date)) |>
  count(year, has_satellite_record) |>
  filter(year >= 2016)
```

## Satellite values and quality information

```{r}
#| label: satellite-summary
integrated |>
  filter(!is.na(modis_date)) |>
  group_by(project_country_id, project_country_name) |>
  summarise(
    years = n(),
    mean_reference_npp = mean(reference_site_npp_kg_c_m2_year, na.rm = TRUE),
    maximum_gap_filled_percent = max(npp_qc_percent, na.rm = TRUE),
    .groups = "drop"
  )
```

The quality percentage remains visible. This introductory workflow does not
silently remove records based on it; a later analysis must justify any
threshold.

## Limitations and appropriate use

- One pixel cannot represent a country, its cropland, or its maize-growing
  areas.
- The reference coordinates are teaching choices, not a probability sample.
- NPP represents total vegetation productivity and is not crop yield.
- A 500 m raster pixel and a country-level agricultural statistic have
  incompatible spatial support even when their identifiers and years join.
- MOD17A3HGF is a processed, gap-filled satellite product rather than an
  untouched sensor observation.
- The satellite snapshot covers only the five evaluation years.

The scientifically responsible conclusion is therefore about the integration
workflow: it demonstrates how to acquire, align, join, audit, and document two
sources. A substantive environmental analysis would require crop masks,
country or production-zone boundaries, area-weighted raster summaries,
growing-season alignment, uncertainty assessment, and a defensible sampling
design.
