---
title: "FAOSTAT–CHIRPS data-integration report"
format: html
execute:
  echo: true
  warning: false
---

This report audits the integration of annual FAOSTAT maize statistics with
country-area CHIRPS precipitation totals for the October-April season ending in
each yield year.

```{r}
source("../scripts/functions.R")
check_required_packages(c("dplyr", "ggplot2", "here", "readr"))
library(dplyr)
library(ggplot2)
library(here)
library(readr)

integrated <- read_csv(
  here("data", "derived", "maize-yield-with-precipitation.csv"),
  show_col_types = FALSE
)
audit <- read_csv(
  here("results", "tables", "data-integration-audit.csv"),
  show_col_types = FALSE
)
```

## Source grains and lineage

- FAOSTAT grain: one project country and calendar year in the prepared panel.
- CHIRPS teaching grain: one project country and October-April season.
- Join key: `project_country_id + year`, where the precipitation year is the
  year in which the season ends.

Provider context is recorded in `metadata/source-metadata.yml`; exact file
checksums and transformations are recorded in `metadata/provenance.yml`.

## Integration audit

```{r}
audit
```

## Coverage

```{r}
integrated |>
  summarise(
    maize_rows = n(),
    precipitation_rows = sum(!is.na(growing_season_precipitation_mm)),
    precipitation_start = min(year[!is.na(growing_season_precipitation_mm)]),
    precipitation_end = max(year[!is.na(growing_season_precipitation_mm)])
  )
```

## Yield and precipitation

```{r}
covered <- integrated |>
  filter(!is.na(growing_season_precipitation_mm))

ggplot(
  covered,
  aes(growing_season_precipitation_mm, yield_tonnes_per_hectare, colour = project_country_name)
) +
  geom_point(size = 2, alpha = 0.8) +
  labs(
    x = "October-April country-area precipitation (mm)",
    y = "Reported maize yield (t/ha)",
    colour = "Country"
  ) +
  theme_minimal()
```

```{r}
covered |>
  group_by(project_country_name) |>
  summarise(
    seasons = n(),
    precipitation_min_mm = min(growing_season_precipitation_mm),
    precipitation_max_mm = max(growing_season_precipitation_mm),
    .groups = "drop"
  )
```

## Interpretation

CHIRPS is satellite-informed and gauge-blended, not a direct rainfall
measurement at every location. Country averages mask subnational conditions
and are not weighted to maize-growing areas. The common October-April season is
a transparent teaching convention rather than a country-specific crop
calendar. A plotted association is descriptive and cannot establish a causal
effect of precipitation on yield.
