---
title: "FAOSTAT teaching extract: data-validation report"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
---

## Purpose and status

This report documents the evidence used to decide whether the fixed FAOSTAT
teaching extract has the expected identity, structure, coverage, and basic
semantic properties. Validation does not prove that provider values are
accurate or that the data are fit for every research question.

```{r}
#| label: setup
library(dplyr)
library(here)
library(readr)

input_file <- here("data", "input", "faostat-maize-yield-sample.csv")
results_file <- here("results", "tables", "data-validation-results.csv")

if (!file.exists(results_file)) {
  stop("Run Rscript scripts/validate-data.R before rendering this report.")
}

maize <- read_csv(input_file, show_col_types = FALSE)
validation <- read_csv(results_file, show_col_types = FALSE)
```

```{r}
#| label: validation-status
validation |>
  count(status, name = "checks")

validation
```

`failure` means the input is not the expected teaching artifact and should not
continue. `warning` preserves a finding that needs interpretation but does not
silently remove or replace observations. `pass` means that the observed data
match one documented expectation.

## Source and provenance

- Artifact: `data/input/faostat-maize-yield-sample.csv`
- Provider: Food and Agriculture Organization of the United Nations (FAO)
- Dataset: FAOSTAT Crops and Livestock Products
- Snapshot created: 3 August 2026
- SHA-256: `fd2c78cae5a5cf2f82d6b6bdc2b3637ce03b597f74e561099f9666af449605be`
- Licence: CC BY 4.0, complemented by the FAO Statistical Database Terms of Use

The complete retrieval history, citation, sharing conditions, and limitations
are recorded in `metadata/provenance.yml`.

## Grain and candidate key

One row represents one maize element for one reporting area and calendar year,
expressed in one unit.

The tested candidate key is:

```text
area + item + element + year + unit
```

The teaching extract uses readable area and item labels because provider codes
were removed when the compact snapshot was created. Future integrations should
use stable provider codes rather than labels wherever those codes are available.

## Dimensions and coverage

```{r}
#| label: dimensions-coverage
tibble(
  rows = nrow(maize),
  columns = ncol(maize),
  countries = n_distinct(maize$area),
  first_year = min(maize$year),
  last_year = max(maize$year)
)

maize |>
  count(area, name = "observations")
```

## Categories, units, and flags

```{r}
#| label: categories-units
maize |>
  count(item, element, unit, name = "observations")

maize |>
  count(flag, name = "observations", sort = TRUE)
```

Flag definitions are recorded in `metadata/flag-code-list.csv`. Flags qualify
the source of a value and must not be discarded without a justified preparation
decision.

## Missingness and ranges

```{r}
#| label: missingness
maize |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "missing")
```

```{r}
#| label: ranges
maize |>
  group_by(element, unit) |>
  summarise(
    minimum = min(value, na.rm = TRUE),
    maximum = max(value, na.rm = TRUE),
    .groups = "drop"
  )
```

A blank source value would be missing, not zero. Plausible ranges depend on the
element and unit; no universal upper threshold is imposed.

## Findings and deferred decisions

The production–area–yield relationship is a diagnostic, not an automatic
repair rule. Differences can arise through rounding, estimation, imputation,
or provider methodology. Any warning remains visible for interpretation during
data preparation.

The following questions remain outside automated validation:

- whether provider estimates accurately represent agricultural conditions;
- whether national statistics are comparable across countries and years;
- whether revisions occurred after the snapshot was created;
- whether country-level annual data are fit for a particular causal question;
- how flags should affect a specific analysis.

No rows are deleted and no source values are changed by validation.
