---
title: "Maize-yield descriptive-data-analysis report"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This report presents the reproducible evidence created by
`scripts/describe-maize-data.R`. It quantifies coverage, distributions, period
differences, temporal dependence, and yield-precipitation association. The
results prepare decisions for later models; they do not estimate causal
effects or prove that a process is stationary.

```{r}
#| label: setup
library(dplyr)
library(here)
library(knitr)
library(readr)

read_result <- function(file_name) {
  path <- here("results", "tables", file_name)
  if (!file.exists(path)) {
    stop(
      "Missing descriptive result: ", path,
      ". Run Rscript scripts/describe-maize-data.R first."
    )
  }
  read_csv(path, show_col_types = FALSE)
}

coverage <- read_result("descriptive-coverage.csv")
yield_summary <- read_result("maize-yield-summary.csv")
yield_period <- read_result("maize-yield-period-summary.csv")
precipitation <- read_result("precipitation-summary.csv")
association <- read_result("yield-precipitation-association.csv")
stationarity <- read_result("stationarity-diagnostic.csv")
```

## Scope and coverage

One observation represents one selected country and year. Yield is a national
FAOSTAT value, while growing-season precipitation is an October-April
country-area CHIRPS estimate. Neither variable represents individual farms or
within-country variation.

```{r}
#| label: coverage
kable(coverage)
```

A `pass` confirms the expected 33 annual values from 1990 through 2022. It does
not establish representativeness beyond the project countries or national
grain.

## Maize-yield distributions

```{r}
#| label: yield-summary
yield_summary |>
  select(
    project_country_name,
    n,
    mean_t_per_ha,
    median_t_per_ha,
    sd_t_per_ha,
    iqr_t_per_ha,
    minimum_t_per_ha,
    maximum_t_per_ha
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable()
```

Mean and median describe location; standard deviation, IQR, and range describe
different aspects of dispersion. Interpret them with
`figures/maize-yield-trends.png`: pooled or full-period statistics can conceal
country trajectories and distribution shape.

## Comparison across periods

The predefined periods are `earlier_history` (1990–2005), `recent_training`
(2006–2017), and `later_test` (2018–2022). They follow the modeling workflow
rather than a contrast selected after examining results.

```{r}
#| label: yield-period-summary
yield_period |>
  select(
    project_country_name,
    analysis_period,
    n,
    mean_t_per_ha,
    median_t_per_ha,
    sd_t_per_ha,
    iqr_t_per_ha
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable()
```

The later test period contains only five values per country. Its location and
dispersion summaries are therefore sensitive to individual years. Period
differences provide evidence about change, not proof for or against
stationarity.

## Growing-season precipitation

```{r}
#| label: precipitation-summary
precipitation |>
  select(
    project_country_name,
    analysis_period,
    n,
    mean_mm,
    median_mm,
    sd_mm,
    iqr_mm
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable()
```

These summaries describe country-area seasonal estimates, not rainfall over
maize-growing land. Spatial aggregation and the common October-April season
limit agronomic interpretation.

## Yield-precipitation association

```{r}
#| label: association-summary
association |>
  filter(scope != "within_country_period") |>
  mutate(
    covariance_mm_t_per_ha = round(covariance_mm_t_per_ha, 3),
    pearson_correlation = round(pearson_correlation, 3)
  ) |>
  kable()
```

Covariance retains both variables' units. Pearson correlation standardizes
linear association. Pooled, within-country, and within-period estimates answer
different questions and can differ because of aggregation, country levels,
trends, and omitted factors. None is a causal effect.

## Evidence relevant to stationarity

```{r}
#| label: stationarity-diagnostic
stationarity |>
  select(
    project_country_name,
    n__recent_training,
    n__later_test,
    mean_change_t_per_ha,
    standardized_mean_change,
    sd_ratio,
    lag_one_pairs,
    lag_one_yield_correlation
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable()
```

The table compares mean and standard deviation between recent training and
later test periods and reports full-series lag-one yield correlation. Review
it with the complete time-series plots and all period summaries. A large
change or correlation is a prompt for investigation; a small value does not
certify stationarity. Formal tests, if added, require their own hypotheses,
trend and lag specifications, and limitations.

## Modeling handoff

```{r}
#| label: modeling-handoff
#| results: asis
handoff_file <- here("results", "descriptive-modeling-handoff.md")
if (!file.exists(handoff_file)) {
  stop("Run Rscript scripts/describe-maize-data.R first.")
}
cat(paste(readLines(handoff_file, warn = FALSE), collapse = "\n"))
```

## Reproducibility statement

The report consumes generated tables rather than recalculating hidden variants
of the analysis. The script uses the validated integrated artifact,
project-relative paths, fixed period definitions, explicit denominators, and
the package versions recorded in `renv.lock`. Generated tables, the handoff,
and rendered HTML are ignored because they can be recreated.
