---
title: "Maize-yield explanatory-analysis report"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This report presents the project's causal question, identification assessment,
regression estimates, diagnostics, and bounded conclusion. The models quantify
associations between prior-year public agricultural expenditure orientation
and subsequent national maize yield. The available observational data do not
identify a causal policy effect.

```{r}
#| label: setup
library(dplyr)
library(here)
library(knitr)
library(readr)

read_result <- function(file_name) {
  path <- here("results", "tables", file_name)
  if (!file.exists(path)) {
    stop(
      "Missing explanatory result: ", path,
      ". Run Rscript scripts/explain-maize-yield.R first."
    )
  }
  read_csv(path, show_col_types = FALSE)
}

sample_audit <- read_result("explanatory-analysis-sample-audit.csv")
support <- read_result("explanatory-exposure-support.csv")
estimates <- read_result("explanatory-model-estimates.csv")
diagnostics <- read_result("explanatory-model-diagnostics.csv")
residual_dependence <- read_result(
  "explanatory-residual-dependence.csv"
)

model_file <- here(
  "results", "models",
  "explanatory-country-year-weather-model.rds"
)
if (!file.exists(model_file)) {
  stop("Run Rscript scripts/explain-maize-yield.R before rendering.")
}
main_model <- readRDS(model_file)
```

## Causal question and estimand

The provisional question asks how subsequent national maize yield would differ
if the prior-year agriculture share of government expenditure were one
percentage point higher among the observed project country-years.

The target contrast can be written as:

$$
E[Y(G + 1\text{ percentage point}) - Y(G)].
$$

The exposure is broad: equal shares can represent different absolute budgets,
sectors, programmes, government levels, and implementation quality. It is not
a randomized treatment or maize-specific expenditure.

## Temporal alignment and sample

Expenditure in year $t-1$ is linked to maize yield in year $t$. This makes
temporal order explicit but does not remove policy-response bias.
The annual expenditure period can overlap the October–April growing season,
so the lag is an alignment convention rather than a precisely timed policy
intervention.

```{r}
#| label: sample-audit
sample_audit |>
  kable()
```

The matched analysis contains 172 observations. Eight countries cover outcome
years 2002–2022, while Zimbabwe contributes four observations because its
expenditure series is available only for 2009–2012. Missing exposure years are
not imputed or treated as zero.

## Causal structure

The tracked causal model uses this simplified diagram:

```text
prior agricultural conditions ─► expenditure share ─► productive capacity ─► maize yield
             │                         ▲                                      ▲
             └─────────────────────────┼──────────────────────────────────────┤
                                       │                                      │
politics, fiscal capacity and crises ──┘                                      │
                                                                              │
weather and precipitation ────────────────────────────────────────────────────┘
```

See `docs/causal-model.md` for variable roles, timing, the full identification
assessment, and the planned adjustment strategy.

## Identification assessment

| Requirement | Project judgment | Consequence |
| --- | --- | --- |
| Consistency | Doubtful | A one-point increase has heterogeneous spending versions |
| Conditional exchangeability | Not established | Policy response and changing common causes remain |
| Positivity | Limited | Country ranges differ and Zimbabwe contributes four observations |
| No interference | Uncertain | Markets, aid, and regional shocks connect countries |
| Measurement validity | Limited | Broad sectors and government-level changes affect comparability |
| Selection and transport | Limited target | Results apply only to observed matched country-years |

Country and year indicators plus precipitation do not establish a sufficient
causal adjustment set.

## Exposure support

The table reports within-country support for the prior-year agriculture share
of government expenditure.

```{r}
#| label: exposure-support
support |>
  filter(scope == "within_country") |>
  select(
    group_name,
    n,
    first_expenditure_year,
    last_expenditure_year,
    minimum_percent,
    median_percent,
    maximum_percent,
    range_percent,
    government_levels
  ) |>
  mutate(across(
    c(minimum_percent, median_percent, maximum_percent, range_percent),
    ~ round(.x, 2)
  )) |>
  kable()
```

Observed support does not prove conditional positivity. Government-level
counts above one also identify possible measurement breaks rather than policy
variation alone.

## Planned specifications

| Model | Added terms | Purpose |
| --- | --- | --- |
| `unadjusted` | None | Pooled association |
| `country` | Country indicators | Account for stable country differences |
| `country_year` | Country and outcome-year indicators | Add common yearly shocks |
| `country_year_weather` | Country, year, and precipitation | Main adjusted association |
| `nonlinear_sensitivity` | Quadratic exposure, country, year, precipitation | Functional-form sensitivity |

Outcome-year precipitation is an environmental cause of yield, not the focal
exposure. Adjusting for it does not resolve omitted fiscal, political, market,
or agricultural causes of expenditure and yield.

## Estimates

```{r}
#| label: model-estimates
estimates |>
  transmute(
    model,
    n,
    estimate = round(
      estimate_t_per_ha_per_percentage_point, 3
    ),
    confidence_low = round(confidence_low, 3),
    confidence_high = round(confidence_high, 3),
    adjusted_r_squared = round(adjusted_r_squared, 3),
    interpretation
  ) |>
  kable()
```

Linear-model estimates represent the modeled change in tonnes per hectare per
one-percentage-point increase in prior-year expenditure share. The nonlinear
sensitivity row reports an average modeled derivative over observed exposure
values.

Differences across specifications are substantive sensitivity evidence. They
are not a search for the model producing the preferred sign or p-value.

## Numerical diagnostics

```{r}
#| label: model-diagnostics
diagnostics |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable()
```

## Main-model diagnostic plots

```{r}
#| label: fig-linear-diagnostics
#| fig-cap: "Standard diagnostic plots for the country-year-weather association model."
#| fig-alt: "Four diagnostic plots showing residuals versus fitted values, a normal quantile comparison, residual scale and leverage for the main linear model."
#| fig-width: 10
#| fig-height: 8
par(mfrow = c(2, 2))
plot(main_model)
par(mfrow = c(1, 1))
```

These plots assess functional form, residual behavior, and influence. They do
not establish exchangeability, consistency, or causal direction.

## Residual temporal dependence

```{r}
#| label: residual-dependence
residual_dependence |>
  filter(model == "country_year_weather") |>
  select(
    project_country_name,
    lag_one_pairs,
    lag_one_residual_correlation
  ) |>
  mutate(
    lag_one_residual_correlation =
      round(lag_one_residual_correlation, 3)
  ) |>
  kable()
```

Lag-one residual correlation indicates remaining temporal structure. Ordinary
regression intervals assume an error structure that may not represent this
dependence. Only nine countries also limit simple cluster-based remedies.

## Bounded conclusion

```{r}
#| label: bounded-conclusion
#| results: asis
conclusion_file <- here(
  "results", "explanatory-modeling-conclusion.md"
)
if (!file.exists(conclusion_file)) {
  stop("Run Rscript scripts/explain-maize-yield.R first.")
}
cat(paste(readLines(conclusion_file, warn = FALSE), collapse = "\n"))
```

## Reproducibility statement

The causal question and assumptions are tracked separately from generated
results. The script checks both prepared inputs, constructs the lag explicitly,
audits matching and selection, uses prespecified models, and writes result
tables before this report renders them. Generated tables, model objects,
conclusions, diagnostic figures, and HTML are ignored because the tracked
inputs, environment, script, causal model, and report source recreate them.
