---
title: "Maize-yield explanatory-modeling report"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This report presents the project's causal question, identification assessment,
regression estimates, diagnostics, and bounded conclusion. The models quantify
precipitation-yield associations under stated specifications. The available
national observational data do not identify the proposed causal effect.

```{r}
#| label: setup
library(dplyr)
library(here)
library(knitr)
library(readr)

read_result <- function(file_name) {
  path <- here("results", "tables", file_name)
  if (!file.exists(path)) {
    stop(
      "Missing explanatory result: ", path,
      ". Run Rscript scripts/explain-maize-yield.R first."
    )
  }
  read_csv(path, show_col_types = FALSE)
}

support <- read_result("explanatory-exposure-support.csv")
estimates <- read_result("explanatory-model-estimates.csv")
diagnostics <- read_result("explanatory-model-diagnostics.csv")
residual_dependence <- read_result(
  "explanatory-residual-dependence.csv"
)

model_file <- here(
  "results", "models", "explanatory-country-time-model.rds"
)
if (!file.exists(model_file)) {
  stop("Run Rscript scripts/explain-maize-yield.R before rendering.")
}
country_time_model <- readRDS(model_file)
```

## Causal question and estimand

The provisional question asks how national maize yield would differ under a
100 mm increase in October-April country-area precipitation among the nine
project countries from 1990 through 2022.

The target contrast can be written as:

$$
E[Y(P + 100\text{ mm}) - Y(P)].
$$

This contrast is ambiguous because a seasonal total hides precipitation
timing, intensity, and spatial distribution. The project records weather
variation, not a controlled rainfall intervention.

## Causal structure

The tracked causal model uses this simplified diagram:

```text
seasonal weather ───► precipitation ───► maize yield
       │                                      ▲
       └──────────────────────────────────────┘

country/time context ─► precipitation
          │                    │
          └───────────────────►yield

irrigation, soils, inputs and management ────►yield
```

See `docs/causal-model.md` for the full variable inventory, arrow reasoning,
adjustment strategy, and identification assessment.

## Identification assessment

| Requirement | Project judgment | Consequence |
| --- | --- | --- |
| Consistency | Doubtful | A 100 mm difference has multiple versions |
| Conditional exchangeability | Not established | Changing weather and agricultural common causes are omitted |
| Positivity | Partly assessable | Some within-country or period contrasts may extrapolate |
| No interference | Uncertain | Regional weather, water, trade, and shocks connect units |
| Measurement | Limited | National yield and country-area rainfall are imperfectly aligned |
| Selection and transport | Limited target | Results do not generalize automatically |

Country and year terms are available proxies. They are not treated as a
sufficient causal adjustment set.

## Exposure support

The following table reports within-country precipitation support in units of
100 mm:

```{r}
#| label: exposure-support
support |>
  filter(scope == "within_country") |>
  select(
    group_name,
    n,
    minimum_100mm,
    q25_100mm,
    median_100mm,
    q75_100mm,
    maximum_100mm,
    range_100mm
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable()
```

Observed variation is necessary but does not prove conditional positivity.
Country-period support contains as few as five observations and cannot justify
arbitrary local contrasts.

## Planned model sequence

| Model | Terms beyond intercept | Intended statistical comparison |
| --- | --- | --- |
| `unadjusted` | Precipitation | Pooled linear association |
| `country` | Precipitation and country indicators | Common within-country association |
| `time` | Precipitation and common linear year | Association conditional on shared trend |
| `country_time` | Precipitation, country, and year | Within-country association around shared trend |
| `nonlinear_sensitivity` | Quadratic precipitation, country, and year | Functional-form sensitivity |

## Model estimates

```{r}
#| label: model-estimates
estimates |>
  select(
    model,
    n,
    estimate_t_per_ha_per_100mm,
    standard_error,
    confidence_low,
    confidence_high,
    r_squared,
    adjusted_r_squared,
    interpretation
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable()
```

Linear-model estimates represent modeled yield differences in tonnes per
hectare per 100 mm. The nonlinear row reports the average modeled derivative
at observed exposure values; the slope itself varies with precipitation.

Intervals use default ordinary-linear-model assumptions. They do not represent
uncertainty from unmeasured confounding, exposure ambiguity, measurement error,
selection, interference, or arbitrary specification.

## Numerical diagnostics

```{r}
#| label: model-diagnostics
diagnostics |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable()
```

Cook's-distance thresholds identify observations for investigation, not
automatic deletion. Residual-fitted correlations summarize selected patterns
but do not replace visual inspection.

## Country-and-time diagnostic plots

```{r}
#| label: fig-linear-diagnostics
#| fig-cap: "Standard diagnostic plots for the country-and-time association model."
#| fig-alt: "Four diagnostic plots showing residuals versus fitted values, a normal quantile comparison, residual scale and leverage for the country-and-time linear model."
#| fig-width: 10
#| fig-height: 8
par(mfrow = c(2, 2))
plot(country_time_model)
par(mfrow = c(1, 1))
```

These plots assess functional form, residual behavior, and influence. They do
not establish exchangeability, consistency, or causal direction.

## Residual temporal dependence

```{r}
#| label: residual-dependence
residual_dependence |>
  filter(model == "country_time") |>
  select(
    project_country_name,
    lag_one_pairs,
    lag_one_residual_correlation
  ) |>
  mutate(
    lag_one_residual_correlation =
      round(lag_one_residual_correlation, 3)
  ) |>
  kable()
```

Lag-one residual correlation indicates remaining temporal structure. Default
ordinary regression intervals assume an error structure that may not represent
this dependence. Nine country clusters also limit simple asymptotic remedies.

## Bounded conclusion

```{r}
#| label: bounded-conclusion
#| results: asis
conclusion_file <- here(
  "results", "explanatory-modeling-conclusion.md"
)
if (!file.exists(conclusion_file)) {
  stop("Run Rscript scripts/explain-maize-yield.R first.")
}
cat(paste(readLines(conclusion_file, warn = FALSE), collapse = "\n"))
```

## Reproducibility statement

The causal question and assumptions are tracked separately from generated
results. The script reads the validated integrated artifact, checks its grain
and completeness, uses fixed model specifications and explicit units, and
writes tables before the report renders them. Generated tables, model objects,
the conclusion, diagnostic figures, and HTML are ignored because the tracked
inputs, environment, script, causal model, and report source recreate them.
