---
title: "Understanding Changes in Maize Yield in Southern Africa"
author: "Learner name"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  echo: true
  warning: true
  message: false
---

## Research question

How have maize yields changed across selected Southern African countries, and how well do simple models predict recent observed yields?

## Data and provenance

The intended source is FAOSTAT Crops and Livestock Products data. During offline use, the project may use a course-provided teaching sample. Confirm the active source before interpreting results.

```{r}
#| label: setup
library(dplyr)
library(ggplot2)
library(here)
library(readr)

maize <- read_csv(
  here("data-processed", "maize-yield-panel.csv"),
  show_col_types = FALSE
)
```

## Data preparation

Describe filtering, duplicate handling, reshaping, unit conversion, missing values, and derived variables. Refer to `scripts/02-prepare-maize-data.R` for the complete transformation.

```{r}
#| label: data-overview
maize |>
  summarise(
    rows = n(),
    countries = n_distinct(country),
    first_year = min(year, na.rm = TRUE),
    last_year = max(year, na.rm = TRUE),
    missing_yield = sum(is.na(yield_tonnes_per_hectare))
  )
```

## Exploratory analysis

```{r}
#| label: fig-yield-trends
#| fig-cap: "Maize yield trends by country."
ggplot(maize, aes(year, yield_tonnes_per_hectare)) +
  geom_line() +
  facet_wrap(vars(country), scales = "free_y") +
  labs(x = NULL, y = "Yield (tonnes per hectare)") +
  theme_minimal()
```

Address:

- Which countries have the highest recent yields?
- How have yields changed over time?
- Are production changes more consistent with yield changes or harvested-area changes?

## Modeling

```{r}
#| label: model-results
performance <- read_csv(
  here("data-processed", "model-performance.csv"),
  show_col_types = FALSE
)
performance
```

Interpret the coefficient on `year` in the descriptive model and compare predictive performance on 2018 onward data. Explain why a model can be useful for description without being suitable for causal claims.

## Limitations

Discuss measurement quality, missingness, country comparability, revisions, omitted variables, the simplicity of the models, and the difference between association and causation.

## Conclusion

Provide two or three findings supported by the analysis. Avoid claims that go beyond the data and methods.

## Reproducibility statement

This report was generated from version-controlled scripts using relative paths. Raw data were not edited manually. Package versions should be recorded with `renv`. The source, download date, transformation steps, and known limitations are documented in `metadata/`.

```{r}
#| label: session-info
sessionInfo()
```
