---
title: "Maize-yield predictive-modeling report"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This report evaluates three predefined benchmark procedures on held-out later
maize-yield observations. It separates training from testing, preserves
row-level predictions, reports complementary errors, and states the limits of
the evidence. It is a teaching benchmark rather than an operational forecast.

```{r}
#| label: setup
library(dplyr)
library(ggplot2)
library(here)
library(knitr)
library(readr)
library(tidyr)

read_result <- function(file_name) {
  path <- here("results", "tables", file_name)
  if (!file.exists(path)) {
    stop(
      "Missing predictive result: ", path,
      ". Run Rscript scripts/model-maize-yield.R first."
    )
  }
  read_csv(path, show_col_types = FALSE)
}

split_audit <- read_result("predictive-split-audit.csv")
predictions <- read_result("maize-yield-predictions.csv")
performance <- read_result("model-performance.csv")
country_performance <- read_result(
  "predictive-performance-by-country.csv"
)
```

## Prediction contract

| Component | Project decision |
| --- | --- |
| Target | Annual national log maize yield |
| Prediction unit | One country-year |
| Population | Nine project countries represented during training |
| Training period | 1990–2017 |
| Test period | 2018–2022 |
| Information | Country identity and calendar year |
| Use | Reproducible later-period teaching benchmark |

The task has no operational issue date or decision-specific loss. It does not
validate a pre-season forecast, unseen-country prediction, or high-stakes use.

## Temporal split audit

```{r}
#| label: split-summary
split_audit |>
  group_by(partition) |>
  summarise(
    countries = n(),
    rows = sum(rows),
    first_year = min(first_year),
    last_year = max(last_year),
    missing_target = sum(missing_target),
    duplicate_country_years = sum(duplicate_country_years),
    .groups = "drop"
  ) |>
  kable()
```

The split learns from earlier years and evaluates later observations. Because
all test countries occur in training, it measures temporal transfer for known
countries rather than geographic transfer to new countries.

## Candidate procedures

| Candidate | Prediction rule | Structure tested |
| --- | --- | --- |
| `historical-mean` | One training-period mean | Constant baseline |
| `linear-trend` | Pooled linear extrapolation | Common temporal change |
| `country-model` | Country indicators and common trend | Persistent country differences plus time |

Every mean and coefficient is estimated from 1990–2017 only. No realized
precipitation is used because the task does not define an issue time at which
that information would be available.

## Overall held-out performance

```{r}
#| label: overall-performance
performance |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable()
```

Mean error uses `observed - predicted`; positive values indicate
underprediction. MAE and RMSE remain in log-yield units. RMSE gives larger
errors more influence than MAE.

## Performance by country

```{r}
#| label: country-performance
country_performance |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable()
```

Each country contributes only five held-out observations. These values reveal
heterogeneity and possible failures but are not precise estimates of future
country performance.

## Observed and predicted test outcomes

```{r}
#| label: fig-predictive-comparison
#| fig-cap: "Observed and predicted log maize yield for held-out years."
#| fig-alt: "Faceted time plots for nine countries comparing observed log maize yield with historical mean, linear trend and country plus time predictions from 2018 through 2022."
#| fig-width: 10
#| fig-height: 8
prediction_long <- predictions |>
  pivot_longer(
    cols = c(historical_mean, linear_trend, country_model),
    names_to = "model",
    values_to = "estimate"
  ) |>
  mutate(
    model = recode(
      model,
      historical_mean = "historical-mean",
      linear_trend = "linear-trend",
      country_model = "country-model"
    )
  )

ggplot(prediction_long, aes(x = year)) +
  geom_line(
    aes(y = estimate, colour = model, group = model),
    linewidth = 0.7
  ) +
  geom_point(aes(y = estimate, colour = model), size = 1.5) +
  geom_line(aes(y = log_yield, group = 1), linewidth = 0.8) +
  geom_point(aes(y = log_yield), size = 1.8) +
  facet_wrap(~ country, scales = "free_y", ncol = 3) +
  scale_x_continuous(breaks = c(2018, 2020, 2022)) +
  labs(
    x = "Year",
    y = "Log yield",
    colour = "Model"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

The figure exposes timing, direction, and subgroup patterns hidden by an
overall average. It should be interpreted together with both error metrics.

## Bounded conclusion

```{r}
#| label: bounded-conclusion
#| results: asis
conclusion_file <- here("results", "predictive-modeling-conclusion.md")
if (!file.exists(conclusion_file)) {
  stop("Run Rscript scripts/model-maize-yield.R before rendering.")
}
cat(paste(readLines(conclusion_file, warn = FALSE), collapse = "\n"))
```

## Reproducibility statement

The tracked script checks the input grain and coverage, enforces a fixed
temporal split, fits candidates on training observations only, preserves
row-level held-out predictions, and generates all metrics, figures, model
objects, and conclusions. Generated artifacts and rendered HTML are ignored
because fixed inputs, the pinned environment, code, report source, and
documentation recreate them.
